Racial and Ethnic Differences in ETI

### 2023-09-15 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti.qmd
race ETI all data manipulation in separate file

```{r data cleaning}
#| include: false
# source("/Users/timvigers/GitHub/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/create_race_eti_analysis_dataset.R")
load("~/Documents/Work/Vigers/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/analysis_dataset.RData")
```

@@

In the tables below, race and ethnicity are considered separately, so some participants may be double counted. For example, a participant identifying as Black and Hispanic would contribute to both categories.

## Race and ethnicity overall

The table below shows the race and ethnicity proportions for the entire cohort included in the CFFPR dataset.

```{r}
table1(~ race | Hispanicrace, data=demo)
```

@@

```{r}
table1(~race+Hispanicrace|factor(ReviewYear),data = annual)
```

## Race percentages overall

Based on the above tables, I decided to combine race and ethnicity with Hispanic "overriding" other racial identities. So if a participant was reported as both Black and Hispanic, they were consider Hispanic for the following analyses. Those with missing data for Hispanic ethnicity were moved to the "Indigenous/Mixed/Other/Unknown Race" category.

```{r}
demo$race <- as.character(demo$race)
demo$race[demo$Hispanicrace == "Yes"] <- "Hispanic or Latino"
demo$race[demo$Hispanicrace == "Unknown"] <- "Hispanic or Latino"
demo$race <- factor(demo$race,
  levels = c(
    "White", "Black or African American", "Hispanic or Latino", "Asian",
    "American Indian or Alaska Native",
    "Native Hawaiian or Other Pacific Islander", "Mixed/Other/Unknown Race"
  ),
  labels = c(
    "White", "Black or African American", "Hispanic or Latino", "Asian",
    "Indigenous/Mixed/Other/Unknown Race",
    "Indigenous/Mixed/Other/Unknown Race", "Indigenous/Mixed/Other/Unknown Race"
  )
)
encounter$race <- NULL
encounter <- left_join(encounter, demo[, c("eDWID", "race")], by = join_by(eDWID))
label(encounter$race) <- "Race/Ethnicity"
label(demo$race) <- "Race/Ethnicity"
```

```{r}
table1(~race,data = demo)
```

@@

@@

```{r}
table1(~Gender+WasGenotyped+F508_num|race,data=demo)
```

@@

```{r}
#| cache: true
by_year = encounter %>%
  group_by(eDWID, reviewyear) %>%
  summarise(across(c(race, Gender, WasGenotyped, F508_num, contains("insurance")), first),
    `Age at Last Visit` = last(na.omit(encounterage)),
    `Mean ppFEV` = mean(GLI_FEV1_pct_predicted, na.rm = T),
    `Weight at Last Visit` = last(na.omit(weight)),
    `Weight %ile at Last Visit` = last(na.omit(weightpercentile)),
    `BMI at Last Visit` = last(na.omit(bmivalue)),
    `BMI %ile at Last Visit` = last(na.omit(bmipercentile)),
    across(contains("Vx"), ~ max(.x)),
    .groups = "drop"
  ) %>%
  select(
    -eDWID, reviewyear, race, Gender, WasGenotyped, F508_num, contains("insurance"),
    `Age at Last Visit`, `Mean ppFEV`, `Weight at Last Visit`,
    `Weight %ile at Last Visit`, `BMI at Last Visit`,
    `BMI %ile at Last Visit`, contains("Vx")
  ) %>%
  rename(any_of(tidy_labels))

```{r}
table1(~Gender+WasGenotyped+F508_num|race,data=by_year)
```

@@

```{r}
encounter %>%
  mutate(eti_elig = factor(eti_elig, ordered = T)) %>%
  group_by(eDWID, reviewyear) %>%
  summarise(
    eti_elig = max(eti_elig), race = first(na.omit(race)),
    .groups = "drop"
  ) %>%
  select(-eDWID) %>%
  rename(any_of(tidy_labels)) %>%
  tbl_summary(by = `Race/Ethnicity`, missing = "Missing") %>%
        bold_labels() %>%
        as_kable_extra()
  tbl_strata(
    strata = Year,
    .tbl_fun =
      ~ .x %>%

  )
```

>>>>

```{r data cleaning}
#| include: false
# source("/Users/timvigers/GitHub/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/create_race_eti_analysis_dataset.R")
load("/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/analysis_dataset.RData")
```

@@

The table below shows the race and ethnicity proportions for the entire cohort included in the CFFPR dataset. I decided to combine race and ethnicity with Hispanic "overriding" other racial identities. So if a participant was reported as both Black and Hispanic, they were consider Hispanic for the following analyses. Those with missing data for Hispanic ethnicity were moved to the "Indigenous/Mixed/Other/Unknown Race" category.

```{r}
table1(~race, data = demo)
```

@@

```{r}
table1(~ race | factor(ReviewYear), data = annual)
```

@@

```{r}
t1_vars <- c(
  "Gender", "WasGenotyped", "F508_num", "mean_ppFEV", "Age_YrEnd",
  "mean_ppFEV", "weight_last", "weight_perc_last", "bmi_last",
  "bmi_perc_last", "Vx770", "Vx809comb", "Vx661comb", "Vx445comb"
)
t1_form <- as.formula(paste("~", paste(t1_vars, collapse = "+"), "|factor(ReviewYear)"))
```

@@

```{r}
table1(~ Gender + WasGenotyped + F508_num | race, data = demo)
```

@@

```{r}
table1(t1_form, data = annual)
```

@@

## Overall by year

```{r}
table1(~ eti_elig | factor(ReviewYear), data = annual, overall = F)
```

## By race and year

```{r}
